\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-15 k-planar.tex
Version submitted to DCG

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
%\usepackage{iwona}
\usepackage{thm-restate}

@@

%\usepackage[mathlines]{lineno}
%% \setlength{\linenumbersep}{2em}
%\linenumbers
%% \rightlinenumbers
%% \linenumbers
%\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
% \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
% \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
% \renewenvironment{#1}%
%    {\linenomath\csname old#1\endcsname}%
%    {\csname oldend#1\endcsname\endlinenomath}}% 
%\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
% \patchAmsMathEnvironmentForLineno{#1}%
% \patchAmsMathEnvironmentForLineno{#1*}}%
%\AtBeginDocument{%
%\patchBothAmsMathEnvironmentsForLineno{equation}%
%\patchBothAmsMathEnvironmentsForLineno{align}%
%\patchBothAmsMathEnvironmentsForLineno{flalign}%
%\patchBothAmsMathEnvironmentsForLineno{alignat}%
%\patchBothAmsMathEnvironmentsForLineno{gather}%
%\patchBothAmsMathEnvironmentsForLineno{multline}%
%}
%
%
%

@@

\setlength{\parskip}{0.85ex}

@@

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs and other more general classes of graphs. 
\notex{PM}{I removed the following sentences because neither Vida nor could find anything about $k$-planar graphs in \cite{FP08} and, in any case, the proof is easy.  This could easily upset a referee.
% \note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }
Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.}
The main results of this paper generalise the following recent theorem of \citet{dujmovic.joret.ea:planar} to a setting that includes $k$-planar graphs.
 % as well as some other interesting non-minor-closed graph families.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$;
\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ such that $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most $4$; 
\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+7$. 
\end{compactenum}
%PM: Keep 'em all: \textcolor{red}{eliminate some of these results? just keep (b)? }
\end{thm}

@@

The definition of $k$-planar graphs naturally generalises for other surfaces. A graph $G$ drawn on a surface $\Sigma$ is $(\Sigma,k)$-plane if every edge of $G$ is involved in at most $k$ crossings.  A graph $G$ is $(g,k)$-planar if it is isomorphic to some $(\Sigma,k)$-plane graph, for some surface $\Sigma$ with Euler genus at most $g$. \citet{grigoriev.bodlaender:algorithms} study approximation algorithms for $(g,k)$-plane graphs.
 %\citet[Theorem~22]{dujmovic.joret.ea:planar} show that every genus-$g$ graph has a layered $H$-partition of layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most 4. 
\cref{AddDummy} immediately generalises as follows:

@@

Prior to this work, the strongest structural description of $k$-planar or $(g,k)$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{T})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{T}$ of $G$. The layered width of $(\mathcal{L},\mathcal{T})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{T}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$. \citet{dujmovic.morin.ea:layered} proved that planar graphs have layered treewidth at most 3, that graphs of Euler genus $g$ have layered treewidth at most $2g+3$, and more generally that a minor-closed class has bounded layered treewidth if and only if it excludes some apex graph. \citet{dujmovic.eppstein.ea:structure} show that every $k$-planar graph has layered treewidth at most $6(k+1)$, and more generally that 
every $(g,k)$-planar graph has layered treewidth at most $(4g+6)(k+1)$.

If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, 
\citet{dujmovic.joret.ea:planar} proved that if a graph $G$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $k$, then $G$ has layered treewidth at most $(k+1)\ell$. 
% So any property that holds for graphs of bounded layered treewidth also holds for $G$. For example, Norin proved that every $n$-vertex graph with layered treewidth at most $t$ has treewidth less than $2\sqrt{tn}$ (see \citep{dujmovic.morin.ea:layered}). Thus, $G$ has treewidth at most $2\sqrt{(k+1)\ell n}$. This in turn leads to a $O(\sqrt{(k+1)\ell n})$ balanced separator theorem for such graphs \citep{RS-II}. So, in the case of $k$-planar graphs, \cref{k-planar} implies an analogue of the separator theorem of \citet{FP08} mentioned in the introduction, albeit with worse dependence on $k$. 
What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.

@@

The following lemma shows how to interpret an $H$-partition of $G$ and a tree-partition of $H$ as a hierarchical decomposition of $G$; refer to \cref{fig:generalized-tripod}.

@@

%%%  Squashing the bibliography 
  \let\oldthebibliography=\thebibliography
  \let\endoldthebibliography=\endthebibliography
  \renewenvironment{thebibliography}[1]{%
    \begin{oldthebibliography}{#1}%
      \setlength{\parskip}{0.25ex}%
      \setlength{\itemsep}{0.25ex}%
  }{\end{oldthebibliography}}

@@

@@

\bibitem[Ajtai et~al.(1982)Ajtai, Chv{\'a}tal, Newborn, and
  Szemer{\'e}di]{ajtai.chvatal.ea:crossing-free}
Mikl{\'o}s Ajtai, Va{\v{s}}ek Chv{\'a}tal, Monroe~M. Newborn, and Endre
  Szemer{\'e}di.
\newblock Crossing-free subgraphs.
\newblock \emph{Ann. Discrete Math.}, 12:\penalty0 9--12, 1982.
\newblock \doi{10.1016/S0304-0208(08)73484-4}.

\bibitem[Alam et~al.(2018)Alam, Bekos, Gronemann, Kaufmann, and
  Pupyrev]{ABGKP18}
Jawaherul~Md. Alam, Michael~A. Bekos, Martin Gronemann, Michael Kaufmann, and
  Sergey Pupyrev.
\newblock Queue layouts of planar 3-trees.
\newblock In Therese~C. Biedl and Andreas Kerren, editors, \emph{Proc. 26th
  International Symposium on Graph Drawing and Network Visualization} ({GD}
  2018), vol.~11282 of \emph{Lecture Notes in Computer Science}, pp.~213--226. Springer, 2018.
\newblock \doi{10.1007/978-3-030-04414-5\_15}.

\bibitem[Alon et~al.(2002)Alon, Grytczuk, Ha{\l}uszczak, and
  Riordan]{AGHR-RSA02}
Noga Alon, {Jaros{\l}aw} Grytczuk, Mariusz Ha{\l}uszczak, and Oliver Riordan.
\newblock Nonrepetitive colorings of graphs.
\newblock \emph{Random Structures Algorithms}, 21\penalty0 (3-4):\penalty0
  336--346, 2002.
\newblock \doi{10.1002/rsa.10057}.

@@

\bibitem[Bonamy et~al.(2019)Bonamy, Gavoille, and
  Pilipczuk]{bonamy.gavoille.ea:shorter}
Marthe Bonamy, Cyril Gavoille, and Michał Pilipczuk.
\newblock Shorter implicit representation for planar graphs.
\newblock In S.~Cabello and B.~Mohar, editors, \emph{9th Slovenian
  International Conference on Graph Theory} (SICGT), page~63, June 2019.

\bibitem[Bose et~al.(2001)Bose, Morin, Stojmenović, and
  Urrutia]{bose.morin.ea:routing}
Prosenjit Bose, Pat Morin, Ivan Stojmenovi\'c, and Jorge Urrutia.
\newblock Routing with guaranteed delivery in ad hoc wireless networks.
\newblock \emph{Wireless Networks}, 7\penalty0 (6):\penalty0 609--616, 2001.
\newblock \doi{10.1023/A:1012319418150}.

\bibitem[Bose et~al.(2012)Bose, Dujmovi\'c, Hurtado, Iacono, Langerman, Meijer,
  Sacrist\'an, Saumell, and Wood]{ProximityGraphs}
Prosenjit Bose, Vida Dujmovi\'c, Ferran Hurtado, John Iacono, Stefan Langerman,
  Henk Meijer, Vera Sacrist\'an, Maria Saumell, and David~R. Wood.
\newblock Proximity graphs: $e$, $\delta$, $\delta$, $\delta$, $\chi$ and
  $\omega$.
\newblock \emph{Internat. J. Comput. Geom.  Appl.},
  22\penalty0 (5):\penalty0 439--469, 2012.

\bibitem[Chen et~al.(2002)Chen, Grigni, and Papadimitriou]{CGP02}
Zhi-Zhong Chen, Michelangelo Grigni, and Christos~H. Papadimitriou.
\newblock Map graphs.
\newblock \emph{J. ACM}, 49\penalty0 (2):\penalty0 127--138, 2002.
\newblock \doi{10.1145/506147.506148}.

\bibitem[D{\k e}bski et~al.(2019)D{\k e}bski, Felsner, Micek, and Schr{\"
  o}der]{micek:personal}
  Micha{\l} D{\k e}bski, Stefan Felsner, Piotr Micek, and Felix Schr{\" o}der. 
\newblock Improved bounds for centered colorings, 2019.
\newblock \href{https://arxiv.org/abs/1907.04586}{arXiv:1907.04586}.

\bibitem[Dujmovi\'{c} et~al.(2016)Dujmovi\'{c}, Joret, Kozik, and Wood]{DJKW16}
Vida Dujmovi\'{c}, Gwena\"{e}l Joret, Jakub Kozik, and David~R. Wood.
\newblock Nonrepetitive colouring via entropy compression.
\newblock \emph{Combinatorica}, 36\penalty0 (6):\penalty0 661--686, 2016.
\newblock \doi{10.1007/s00493-015-3070-6}.

\bibitem[Dujmovi\'c et~al.(2017{\natexlab{a}})Dujmovi\'c, Eppstein, and
  Wood]{dujmovic.eppstein.ea:structure}
Vida Dujmovi\'c, David Eppstein, and David~R. Wood.
\newblock Structure of graphs with locally restricted crossings.
\newblock \emph{{SIAM} J. Discrete Math.}, 31\penalty0 (2):\penalty0 805--824,
  2017{\natexlab{a}}.
\newblock \doi{10.1137/16M1062879}.

\bibitem[Dujmovi\'c et~al.(2017{\natexlab{b}})Dujmovi\'c, Morin, and
  Wood]{dujmovic.morin.ea:layered}
Vida Dujmovi\'c, Pat Morin, and David~R. Wood.
\newblock Layered separators in minor-closed graph classes with applications.
\newblock \emph{J. Comb. Theory, Ser. {B}}, 127:\penalty0 111--147,
  2017{\natexlab{b}}.
\newblock \doi{10.1016/j.jctb.2017.05.006}.

@@

\bibitem[Dujmovi\'c et~al.(2019{\natexlab{b}})Dujmovi\'c, Joret, Micek, Morin,
  Ueckerdt, and Wood]{dujmovic.joret.ea:planar}
Vida Dujmovi\'c, Gwena{\"{e}}l Joret, Piotr Micek, Pat Morin, Torsten Ueckerdt,
  and David~R. Wood.
\newblock Planar graphs have bounded queue-number.
\newblock \emph{Proc. 60th Annual Symp. on Foundations of Computer 
  Science} (FOCS '19), to appear. 
  \newblock  \href{http://arxiv.org/abs/1904.04791}{arXiv:1904.04791}, 2019{\natexlab{b}}.

\bibitem[Fomin et~al.(2012)Fomin, Lokshtanov, and Saurabh]{FLS-SODA12}
Fedor~V. Fomin, Daniel Lokshtanov, and Saket Saurabh.
\newblock Bidimensionality and geometric graphs.
\newblock In \emph{Proc. 23rd {A}nnual {ACM}-{SIAM} {S}ymposium on {D}iscrete
  {A}lgorithms} (SODA 2012), pp.~1563--1575, 2012.
\newblock \doi{10.1137/1.9781611973099.124}.

\bibitem[Fox and Pach(2010)]{FP10}
Jacob Fox and J{\'a}nos Pach.
\newblock A separator theorem for string graphs and its applications.
\newblock \emph{Combin. Probab. Comput.}, 19\penalty0 (3):\penalty0 371--390,
  2010.
\newblock \doi{10.1017/S0963548309990459}.

\bibitem[Fox and Pach(2014)]{FP14}
Jacob Fox and J{\'a}nos Pach.
\newblock Applications of a new separator theorem for string graphs.
\newblock \emph{Combin. Probab. Comput.}, 23\penalty0 (1):\penalty0 66--74,
  2014.
\newblock \doi{10.1017/S0963548313000412}.

\bibitem[Grigoriev and Bodlaender(2007)]{grigoriev.bodlaender:algorithms}
Alexander Grigoriev and Hans~L. Bodlaender.
\newblock Algorithms for graphs embeddable with few crossings per edge.
\newblock \emph{Algorithmica}, 49\penalty0 (1):\penalty0 1--11, 2007.
\newblock \doi{10.1007/s00453-007-0010-x}.

\bibitem[Harvey and Wood(2017)]{HW17}
Daniel~J. Harvey and David~R. Wood.
\newblock Parameters tied to treewidth.
\newblock \emph{J. Graph Theory}, 84\penalty0 (4):\penalty0 364--385, 2017.
\newblock \doi{10.1002/jgt.22030}.

\bibitem[Heath et~al.(1992)Heath, Leighton, and Rosenberg]{HLR92}
Lenwood~S. Heath, F.~Thomson Leighton, and Arnold~L. Rosenberg.
\newblock Comparing queues and stacks as mechanisms for laying out graphs.
\newblock \emph{SIAM J. Discrete Math.}, 5\penalty0 (3):\penalty0 398--412,
  1992.
\newblock \doi{10.1137/0405031}.

\bibitem[van~den Heuvel and Wood(2017)]{vdHW17}
Jan van~den Heuvel and David~R. Wood.
\newblock Improper colourings inspired by {H}adwiger's conjecture.
\newblock \emph{J. London Math. Soc.}, 98(1):129--148, 2018. 
\newblock  \doi{10.1112/jlms.12127}. See   \href{https://arxiv.org/abs/1704.06536}{arXiv:1704.06536} for the cited results.

\bibitem[Kannan et~al.(1988)Kannan, Naor, and
  Rudich]{kannan.naor.ea:implicit-stoc}
Sampath Kannan, Moni Naor, and Steven Rudich.
\newblock Implicit representation of graphs.
\newblock In Janos Simon, editor, \emph{Proc.\ 20th Annual {ACM} Symposium on
  Theory of Computing} (STOC '88), pp.~334--343. {ACM}, 1988.
\newblock \doi{10.1145/62212.62244}.

>>>>

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}

@@

\usepackage[mathlines]{lineno}
% \setlength{\linenumbersep}{2em}
\linenumbers
% \rightlinenumbers
% \linenumbers
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
    {\linenomath\csname old#1\endcsname}%
    {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

@@

\setlength{\parskip}{0.82ex}

@@

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs and other more general classes of graphs. 
\notex{PM}{I removed the following sentences because neither Vida nor could find anything about $k$-planar graphs in \cite{FP08} and, in any case, the proof is easy.  This could easily upset a referee.
% \note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }
Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.}

Our main results generalise the following recent theorem of \citet{dujmovic.joret.ea:planar} to a setting that includes $k$-planar graphs.
 % as well as some other interesting non-minor-closed graph families.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$.
%\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ such that $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most $4$.
%\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+8$. 
\end{compactenum}
%PM: Keep 'em all: \textcolor{red}{eliminate some of these results? just keep (b)? }
\end{thm}

@@

The definition of $k$-planar graphs naturally generalises for other surfaces. A graph $G$ drawn on a surface $\Sigma$ is $(\Sigma,k)$-plane if every edge of $G$ is involved in at most $k$ crossings.  A graph $G$ is $(g,k)$-planar if it is isomorphic to some $(\Sigma,k)$-plane graph, for some surface $\Sigma$ with Euler genus at most $g$. 
 %\citet[Theorem~22]{dujmovic.joret.ea:planar} show that every genus-$g$ graph has a layered $H$-partition of layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most 4. 
\cref{AddDummy} immediately generalises as follows:

@@

Prior to this work, the strongest structural description of $k$-planar or $(g,k)$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{T})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{T}$ of $G$. The layered width of $(\mathcal{L},\mathcal{T})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{T}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$. \citet{dujmovic.morin.ea:layered} proved that planar graphs have layered treewidth at most 3, that graphs of Euler genus $g$ have layered treewidth at most $2g+3$, and more generally that a minor-closed class has bounded layered treewidth if and only if it excludes some apex graph. \citet{dujmovic.eppstein.ea:structure} show that every $k$-planar graph has layered treewidth at most $6(k+1)$, and more generally that every $(g,k)$-planar graph has layered treewidth at most $(4g+6)(k+1)$. It follows from this result that $(g,k)$-planar graphs have treewidth $O(\sqrt{(g+1)(k+1)n})$ and thus have balanced separators of the same order, which can also be concluded from the work of 
\citet{FP08}. In related work, \citet{grigoriev.bodlaender:algorithms} used structural results to obtain approximation algorithms for $(g,k)$-planar graphs, and \citet{PachToth97} determined the maximum number of edges in a $k$-planar graph (up to a constant factor).

If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, 
\citet{dujmovic.joret.ea:planar} proved that if a graph $G$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $k$, then $G$ has layered treewidth at most $(k+1)\ell$.  So any property that holds for graphs of bounded layered treewidth also holds for $G$. 
%For example, Norin proved that every $n$-vertex graph with layered treewidth at most $t$ has treewidth less than $2\sqrt{tn}$ (see \citep{dujmovic.morin.ea:layered}). Thus, $G$ has treewidth at most $2\sqrt{(k+1)\ell n}$. This in turn leads to a $O(\sqrt{(k+1)\ell n})$ balanced separator theorem for such graphs \citep{RS-II}. So, in the case of $k$-planar graphs, \cref{k-planar} implies an analogue of the separator theorem of \citet{FP08} mentioned in the introduction, albeit with worse dependence on $k$. 
What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.

@@

The following lemma shows how to interpret an $H$-partition of $G$ and a tree-decomposition of $H$ as a hierarchical decomposition of $G$; refer to \cref{fig:generalized-tripod}.

@@

%%%  Squashing the bibliography 
  \let\oldthebibliography=\thebibliography
  \let\endoldthebibliography=\endthebibliography
  \renewenvironment{thebibliography}[1]{%
    \begin{oldthebibliography}{#1}%
      \setlength{\parskip}{0.4ex}%
      \setlength{\itemsep}{0.4ex}%
  }{\end{oldthebibliography}}

@@

\fontsize{11}{11.75}\selectfont

@@

\bibitem[Ajtai et~al.(1982)Ajtai, Chv{\'a}tal, Newborn, and
  Szemer{\'e}di]{ajtai.chvatal.ea:crossing-free}
Mikl{\'o}s Ajtai, Va{\v{s}}ek Chv{\'a}tal, Monroe~M. Newborn, Endre
  Szemer{\'e}di.
\newblock Crossing-free subgraphs.
\newblock \emph{Ann. Discrete Math.}, 12:\penalty0 9--12, 1982.
\newblock \doi{10.1016/S0304-0208(08)73484-4}.

\bibitem[Alam et~al.(2018)Alam, Bekos, Gronemann, Kaufmann, and
  Pupyrev]{ABGKP18}
Jawaherul~Md. Alam, Michael~A. Bekos, Martin Gronemann, Michael Kaufmann, Sergey Pupyrev.
\newblock Queue layouts of planar 3-trees.
\newblock In Therese~C. Biedl and Andreas Kerren, editors, \emph{Proc. 26th
  International Symposium on Graph Drawing and Network Visualization} (GD '18), vol.~11282 of \emph{Lecture Notes in Computer Science}, pp.~213--226. Springer, 2018.
\newblock \doi{10.1007/978-3-030-04414-5\_15}.

\bibitem[Alon et~al.(2002)Alon, Grytczuk, Ha{\l}uszczak, and
  Riordan]{AGHR-RSA02}
Noga Alon, {Jaros{\l}aw} Grytczuk, Mariusz Ha{\l}uszczak, Oliver Riordan.
\newblock Nonrepetitive colorings of graphs.
\newblock \emph{Random Structures Algorithms}, 21\penalty0 (3-4):\penalty0
  336--346, 2002.
\newblock \doi{10.1002/rsa.10057}.

@@

\bibitem[Bonamy et~al.(2019)Bonamy, Gavoille, and
  Pilipczuk]{bonamy.gavoille.ea:shorter}
Marthe Bonamy, Cyril Gavoille, and Micha{\l} Pilipczuk.
\newblock Shorter implicit representation for planar graphs.
\newblock In S.~Cabello and B.~Mohar, editors, \emph{9th Slovenian
  International Conference on Graph Theory} (SICGT), p.~63, June 2019.

\bibitem[Bose et~al.(2001)Bose, Morin, Stojmenovi\'c, and
  Urrutia]{bose.morin.ea:routing}
Prosenjit Bose, Pat Morin, Ivan Stojmenovi\'c, Jorge Urrutia.
\newblock Routing with guaranteed delivery in ad hoc wireless networks.
\newblock \emph{Wireless Networks}, 7\penalty0 (6):\penalty0 609--616, 2001.
\newblock \doi{10.1023/A:1012319418150}.

\bibitem[Bose et~al.(2012)Bose, Dujmovi\'c, Hurtado, Iacono, Langerman, Meijer,
  Sacrist\'an, Saumell, and Wood]{ProximityGraphs}
Prosenjit Bose, Vida Dujmovi\'c, Ferran Hurtado, John Iacono, Stefan Langerman,
  Henk Meijer, Vera Sacrist\'an, Maria Saumell, David~R. Wood.
\newblock Proximity graphs: $e$, $\delta$, $\delta$, $\delta$, $\chi$ and
  $\omega$.
\newblock \emph{Internat. J. Comput. Geom.  Appl.},
  22\penalty0 (5):\penalty0 439--469, 2012.

\bibitem[Chen et~al.(2002)Chen, Grigni, and Papadimitriou]{CGP02}
Zhi-Zhong Chen, Michelangelo Grigni, Christos~H. Papadimitriou.
\newblock Map graphs.
\newblock \emph{J. ACM}, 49\penalty0 (2):\penalty0 127--138, 2002.
\newblock \doi{10.1145/506147.506148}.

\bibitem[D{\k e}bski et~al.(2019)D{\k e}bski, Felsner, Micek, and Schr{\"o}der]{micek:personal}
  Micha{\l} D{\k e}bski, Stefan Felsner, Piotr Micek, Felix Schr{\" o}der. 
\newblock Improved bounds for centered colorings, 2019.
\newblock \href{https://arxiv.org/abs/1907.04586}{arXiv:1907.04586}.

\bibitem[Dujmovi\'{c} et~al.(2016)Dujmovi\'{c}, Joret, Kozik, and Wood]{DJKW16}
Vida Dujmovi\'{c}, Gwena\"{e}l Joret, Jakub Kozik, David~R. Wood.
\newblock Nonrepetitive colouring via entropy compression.
\newblock \emph{Combinatorica}, 36\penalty0 (6):\penalty0 661--686, 2016.
\newblock \doi{10.1007/s00493-015-3070-6}.

\bibitem[Dujmovi\'c et~al.(2017{\natexlab{a}})Dujmovi\'c, Eppstein, and
  Wood]{dujmovic.eppstein.ea:structure}
Vida Dujmovi\'c, David Eppstein, David~R. Wood.
\newblock Structure of graphs with locally restricted crossings.
\newblock \emph{{SIAM} J. Discrete Math.}, 31\penalty0 (2):\penalty0 805--824,
  2017{\natexlab{a}}.
\newblock \doi{10.1137/16M1062879}.

\bibitem[Dujmovi\'c et~al.(2017{\natexlab{b}})Dujmovi\'c, Morin, and
  Wood]{dujmovic.morin.ea:layered}
Vida Dujmovi\'c, Pat Morin, David~R. Wood.
\newblock Layered separators in minor-closed graph classes with applications.
\newblock \emph{J. Comb. Theory, Ser. {B}}, 127:\penalty0 111--147,
  2017{\natexlab{b}}.
\newblock \doi{10.1016/j.jctb.2017.05.006}.

@@

\bibitem[Dujmovi\'c et~al.(2019{\natexlab{b}})Dujmovi\'c, Joret, Micek, Morin,
  Ueckerdt, and Wood]{dujmovic.joret.ea:planar}
Vida Dujmovi\'c, Gwena{\"{e}}l Joret, Piotr Micek, Pat Morin, Torsten Ueckerdt, David~R. Wood.
\newblock Planar graphs have bounded queue-number.
\newblock \emph{Proc. 60th Annual Symp. on Foundations of Computer 
  Science} (FOCS '19), to appear. 
  \newblock  \href{http://arxiv.org/abs/1904.04791}{arXiv:1904.04791}, 2019{\natexlab{b}}.

\bibitem[Fomin et~al.(2012)Fomin, Lokshtanov, and Saurabh]{FLS-SODA12}
Fedor~V. Fomin, Daniel Lokshtanov, Saket Saurabh.
\newblock Bidimensionality and geometric graphs.
\newblock In \emph{Proc. 23rd {A}nnual {ACM}-{SIAM} {S}ymposium on {D}iscrete
  {A}lgorithms} (SODA 2012), pp.~1563--1575, 2012.
\newblock \doi{10.1137/1.9781611973099.124}.

\bibitem[Fox and Pach(2008)]{FP08}
Jacob Fox, J{\'a}nos Pach.
\newblock Separator theorems and {T}ur\'an-type results for planar intersection graphs.
\newblock \emph{Adv. Math.}, 219(3):1070--1080, 2008.
\newblock \doi{10.1016/j.aim.2008.06.002}.

\bibitem[Fox and Pach(2010)]{FP10}
Jacob Fox, J{\'a}nos Pach.
\newblock A separator theorem for string graphs and its applications.
\newblock \emph{Combin. Probab. Comput.}, 19\penalty0 (3):\penalty0 371--390,
  2010.
\newblock \doi{10.1017/S0963548309990459}.

\bibitem[Fox and Pach(2014)]{FP14}
Jacob Fox, J{\'a}nos Pach.
\newblock Applications of a new separator theorem for string graphs.
\newblock \emph{Combin. Probab. Comput.}, 23\penalty0 (1):\penalty0 66--74,
  2014.
\newblock \doi{10.1017/S0963548313000412}.

\bibitem[Grigoriev and Bodlaender(2007)]{grigoriev.bodlaender:algorithms}
Alexander Grigoriev, Hans~L. Bodlaender.
\newblock Algorithms for graphs embeddable with few crossings per edge.
\newblock \emph{Algorithmica}, 49\penalty0 (1):\penalty0 1--11, 2007.
\newblock \doi{10.1007/s00453-007-0010-x}.

\bibitem[Harvey and Wood(2017)]{HW17}
Daniel~J. Harvey, David~R. Wood.
\newblock Parameters tied to treewidth.
\newblock \emph{J. Graph Theory}, 84\penalty0 (4):\penalty0 364--385, 2017.
\newblock \doi{10.1002/jgt.22030}.

\bibitem[Heath et~al.(1992)Heath, Leighton, and Rosenberg]{HLR92}
Lenwood~S. Heath, F.~Thomson Leighton, Arnold~L. Rosenberg.
\newblock Comparing queues and stacks as mechanisms for laying out graphs.
\newblock \emph{SIAM J. Discrete Math.}, 5\penalty0 (3):\penalty0 398--412,
  1992.
\newblock \doi{10.1137/0405031}.

\bibitem[van~den Heuvel and Wood(2017)]{vdHW17}
Jan van~den Heuvel, David~R. Wood.
\newblock Improper colourings inspired by {H}adwiger's conjecture.
\newblock \emph{J. London Math. Soc.}, 98(1):129--148, 2018. 
\newblock  \doi{10.1112/jlms.12127}. See   \href{https://arxiv.org/abs/1704.06536}{arXiv:1704.06536} for the cited results.
